---
title: 'Report: Plot markers upon integrated Cre3 and Lox2 scRNA samples'
author: "Jingtao Lilue and António Sousa - UBI-IGC (e-mail: <agsousa@igc.gulbenkian.pt>)"
date: "`r format(Sys.Date(), '%d/%m/%Y')`"
output: 
  html_document:
    toc: true 
    toc_float: true
    theme: united
    code_folding: hide
bibliography: bibliography.bib
---

<br>

---

<br>

<br>

## Request

<br>

This report includes a description and the results of the analyses requested by Elisa Jentho at 01/06/2021: 

   1. Plot marker genes upon integrated Cre3 and Lox2 *Plasmodium chabaudi* scRNA-seq samples with `Seurat`: 
   
      + **Markers_for_UMAP.xlsx**: plot the list of markers in the overlapped UMAP of the integrated Cre3 and Lox2 samples.
      
      + **More_markers.xlsx**: plot the list of markers separately across the two UMAPs of the integrated Cre3 and Lox2 samples. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	fig.align = "center",
	message = FALSE,
	warning = FALSE
)
```

```{r klippy, echo=FALSE}
klippy::klippy(position = c('top', 'right'))
```

<br>

<br>

---

<br>

<br>

## About this report

<br>

The R version used was `r paste0(version$major, ".", version$minor)` [@r]. This report was built with the `rmarkdown` R package (v.`r packageVersion("rmarkdown")`) [@rmarkdown_1; @rmarkdown_2; @rmarkdown_3]. See the full list of packages and versions used at the end of this report - **R packages used and respective versions**.

The integrated 10x Lox2 and Cre3 samples processed previously were imported to plot the markers upon the UMAPs with the `Seurat` R package (v.`r packageVersion("Seurat")`) [@seurat_1; @seurat_2; @seurat_3; @seurat_4].

The figures and tables displayed through the report can be download in *pdf* and *tab-separated* format by clicking on the bottom left **Download plot** and **Download table** button that appears after each plot and table, respectively. 

```{r seed}

## Set seed 
set.seed(seed = 1024) # to keep reproducibility

```

```{r import packages, message=FALSE, warning=FALSE}

## Import packages 

library("dplyr", quietly = TRUE)
library("DT", quietly = TRUE) # package to print nice in the html report 
library("Seurat", quietly = TRUE)
library("gplots", quietly = TRUE)
library("Matrix", quietly = TRUE)
library("ggplot2", quietly = TRUE)
library("tidyr", quietly = TRUE)
library("readxl", quietly = TRUE)

```

<br>

<br>

---

<br>

<br>

## Data analysis

<br>

### Import data: Cre3 & Lox2 integrated samples

<br>

```{r import seu int objs}

## Import Seurat object with the re3 & Lox2 integrated samples

# import obj integrated
seu <- readRDS(file = "../results/int/R_objects/seu.rds")

```

<br>

<br>

### Import data: Marker lists

<br>

```{r import mkr lists}

## Import marker lists 

# marker lists to import 
markers2import <- list.files(path = "../data/markers_to_plot", full.names = TRUE)
names(markers2import) <- c("int", "sep") 
# "int": integrated/overlapped UMAP marker list | "sep": separated UMAP marker list

# import lists
markers <- list(
  "int" = read_excel(path = markers2import[1], col_names = TRUE), 
  "sep" = read_excel(path = markers2import[2], col_names = FALSE)
)

# parse lists
markers$int = markers$int %>% select_if(is.character)
colnames(markers$sep) <- "Genes"

```

The two lists of markers can be found below. 

<br>

```{r echo=FALSE}
datatable(data = markers$int, extensions = "Buttons", 
          caption = "Table with the list of markers to plot/highlight in the overlapped UMAP of the integrated Cre3 and Lox2 samples. Imported from the excel file provided: Markers_for_UMAP.xlsx.", 
          options = list(dom = "Blfrtip", buttons = c('csv', 'excel')))     
```

<br>

```{r echo=FALSE}
datatable(data = markers$sep, extensions = "Buttons", 
          caption = "Table with the list of markers to plot/highlight separately across the two UMAPs of the integrated Cre3 and Lox2 samples. Imported from the excel file provided: More_markers.xlsx.", 
          options = list(dom = "Blfrtip", buttons = c('csv', 'excel')))     
```

<br>

<br>

---

<br>

<br>

### Plot markers by cluster

<br>

To plot markers it was used the integrated `Seurat` object with the Cre3 and Lox2 samples. The assay used was `RNA` and the slot `data` with the counts normalized with the `LogNormalize` method ([read more](https://satijalab.org/seurat/reference/normalizedata)). 

The individual plots of the cluster markers highlighted through the integrated UMAPs can be found at: `results/markers/int/` (with each cluster folder holding their respective markers).  

```{r plot int marker list}

## Plot list of markers by cluster

# folder to save results
plot_markers <- "../results/markers"
if ( ! dir.exists(plot_markers) ) dir.create(plot_markers)

# list to save plots
marker_plots[["sep"]] <- marker_plots[["int"]] <- marker_plots <- list()

# check assay 
stopifnot(DefaultAssay(seu) == "RNA")

for ( clt in colnames(markers$int) ) { # loop over cols/cluster markers
  clt_name <- gsub(pattern = " ", replacement = "-", x = clt)
  
  # create new folder for the cluster
  plot_clt_folder <- paste(plot_markers, paste0("int/", clt_name), sep = "/")
  if ( ! dir.exists(plot_clt_folder) ) dir.create(plot_clt_folder, recursive = TRUE)
  
  marker_genes <- markers$int[,clt,drop=TRUE] # retrieve the marker genes for each col/cluster
  
  # save to list
  marker_plots$int[[clt_name]] <- list()
  
  for ( mkr in marker_genes ) { # loop over marker genes within a cluster
    if( ! is.na(mkr) ) { # plot them only if are not 'NA'
      stopifnot( mkr %in% row.names(seu) ) # stop if gene does not exist in the Seurat obj
      marker_plots$int[[clt_name]][[mkr]] <- FeaturePlot(object = seu, features = mkr, 
                                                         slot = "data", reduction = "umap", 
                                                         pt.size = 0.001)
      ggsave(filename = paste(plot_clt_folder, paste0(clt_name, "_", mkr, "_gene_UMAP.pdf"), sep = "/"), 
             plot = marker_plots$int[[clt_name]][[mkr]], height = 5, width = 5)
    }
  }
}

```

<br>

<br>

   + **Cluster 0**
   
<br>

```{r cluster 0, fig.width=10, fig.height=6}
cowplot::plot_grid(plotlist = marker_plots$int$`Cluster-0`)
```

<br>

<br>

   + **Cluster 1**
   
<br>

```{r cluster 1, fig.width=10, fig.height=6}
cowplot::plot_grid(plotlist = marker_plots$int$`Cluster-1`)
```

<br>

<br>

   + **Cluster 2**
   
<br>

```{r cluster 2, fig.width=8, fig.height=7}
cowplot::plot_grid(plotlist = marker_plots$int$`Cluster-2`)
```

<br>

<br>

   + **Cluster 3**
   
<br>

```{r cluster 3, fig.width=10, fig.height=6}
cowplot::plot_grid(plotlist = marker_plots$int$`Cluster-3`)
```

<br>

<br>

   + **Cluster 4**
   
<br>

```{r cluster 4, fig.width=10, fig.height=6}
cowplot::plot_grid(plotlist = marker_plots$int$`Cluster-4`)
```

<br>

<br>

   + **Cluster 5**
   
<br>

```{r cluster 5, fig.width=10, fig.height=6}
cowplot::plot_grid(plotlist = marker_plots$int$`Cluster-5`)
```

<br>

<br>

   + **Cluster 6**
   
<br>

```{r cluster 6, fig.width=10, fig.height=6}
cowplot::plot_grid(plotlist = marker_plots$int$`Cluster-6`)
```

<br>

<br>

   + **Cluster 7**
   
<br>

```{r cluster 7, fig.width=10, fig.height=6}
cowplot::plot_grid(plotlist = marker_plots$int$`Cluster-7`)
```

<br>

<br>

   + **Cluster 8**
   
<br>

```{r cluster 8, fig.width=10, fig.height=6}
cowplot::plot_grid(plotlist = marker_plots$int$`Cluster-8`)
```

<br>

<br>

---

<br>

<br>

### Plot other markers split by sample

<br>

To plot markers it was used the integrated `Seurat` object with the Cre3 and Lox2 samples. The assay used was `RNA` and the slot `data` with the counts normalized with the `LogNormalize` method ([read more](https://satijalab.org/seurat/reference/normalizedata)). 

The individual plots of the markers split by sample highlighted through the integrated UMAPs can be found at: `results/markers/sep/`.  

>WARNING: The genes PCHAS-051500 and PCHAS-141240 provided as markers to plot were not found across the scRNA-seq data sets, and, thus they were not plotted.  

```{r plot sep marker list}

## Plot list of markers by sample

# folder to save results
plot_sep_markers <- "../results/markers/sep"
if ( ! dir.exists(plot_sep_markers) ) dir.create(plot_sep_markers)

# Seurat: split objects by sample
sobjList <- SplitObject(seu, split.by = "orig.ident") 

# check assay 
stopifnot(DefaultAssay(seu) == "RNA")
stopifnot( c(DefaultAssay(sobjList$Cre3), DefaultAssay(sobjList$Lox2)) == "RNA" )
for ( mkr in markers$sep$Genes ) { # loop over marker genes within a cluster
  #stopifnot( mkr %in% row.names(seu) ) # stop if gene does not exist in the Seurat obj
  if ( mkr %in% row.names(seu) ) {
    marker_plots$sep[[mkr]] <- list()
    for ( sobj in rev(names(sobjList)) ) {
            marker_plots$sep[[mkr]][[sobj]] <- FeaturePlot(object = sobjList[[sobj]], features = mkr,
                                                           slot = "data", reduction = "umap", 
                                                           pt.size = 0.001)
            ggsave(filename = paste(plot_sep_markers, paste0(sobj, "_", mkr, "_gene_UMAP_split_by_sample.pdf"), sep = "/"), 
                   plot = marker_plots$sep[[mkr]][[sobj]], height = 5, width = 5)
    }
  }
}

```

<br>

<br>

   + **Marker PCHAS-021240**
   
<br>

```{r PCHAS-021240, fig.width=10, fig.height=5}
cowplot::plot_grid(plotlist = marker_plots$sep$`PCHAS-021240`)
```

<br>

<br>

   + **Marker PCHAS-121330**
   
<br>

```{r PCHAS-121330, fig.width=10, fig.height=5}
cowplot::plot_grid(plotlist = marker_plots$sep$`PCHAS-121330`)
```

<br>

<br>

   + **Marker PCHAS-100390**
   
<br>

```{r PCHAS-100390, fig.width=10, fig.height=5}
cowplot::plot_grid(plotlist = marker_plots$sep$`PCHAS-100390`)
```

<br>

<br>

   + **Marker PCHAS-140270**
   
<br>

```{r PCHAS-140270, fig.width=10, fig.height=5}
cowplot::plot_grid(plotlist = marker_plots$sep$`PCHAS-140270`)
```

<br>

<br>

   + **Marker PCHAS-143360**
   
<br>

```{r PCHAS-143360, fig.width=10, fig.height=5}
cowplot::plot_grid(plotlist = marker_plots$sep$`PCHAS-143360`)
```

<br>

<br>

   + **Marker PCHAS-110810**
   
<br>

```{r PCHAS-110810, fig.width=10, fig.height=5}
cowplot::plot_grid(plotlist = marker_plots$sep$`PCHAS-110810`)
```

<br>

<br>

   + **Marker PCHAS-040720**
   
<br>

```{r PCHAS-040720, fig.width=10, fig.height=5}
cowplot::plot_grid(plotlist = marker_plots$sep$`PCHAS-040720`)
```

<br>

<br>

   + **Marker PCHAS-031680**
   
<br>

```{r PCHAS-031680, fig.width=10, fig.height=5}
cowplot::plot_grid(plotlist = marker_plots$sep$`PCHAS-031680`)
```

<br>

<br>

   + **Marker PCHAS-122070**
   
<br>

```{r PCHAS-122070, fig.width=10, fig.height=5}
cowplot::plot_grid(plotlist = marker_plots$sep$`PCHAS-122070`)
```

<br>

<br>

   + **Marker PCHAS-142660**
   
<br>

```{r PCHAS-142660, fig.width=10, fig.height=5}
cowplot::plot_grid(plotlist = marker_plots$sep$`PCHAS-142660`)
```

<br>

<br>

   + **Marker PCHAS-146340**
   
<br>

```{r PCHAS-146340, fig.width=10, fig.height=5}
cowplot::plot_grid(plotlist = marker_plots$sep$`PCHAS-146340`)
```

<br>

<br>

   + **Marker PCHAS-072660**
   
<br>

```{r PCHAS-072660, fig.width=10, fig.height=5}
cowplot::plot_grid(plotlist = marker_plots$sep$`PCHAS-072660`)
```

<br>

<br>

   + **Marker PCHAS-081890**
   
<br>

```{r PCHAS-081890, fig.width=10, fig.height=5}
cowplot::plot_grid(plotlist = marker_plots$sep$`PCHAS-081890`)
```

<br>

<br>

   + **Marker PCHAS-113160**
   
<br>

```{r PCHAS-113160, fig.width=10, fig.height=5}
cowplot::plot_grid(plotlist = marker_plots$sep$`PCHAS-113160`)
```

<br>

<br>

---

<br>

<br>
    
## Deliver

Folders inside the project folder:

   + results: folder that contains all the results obtained in this analysis;

   + report: folder that contains the report and code used herein;

   + data: folder that contains the data used herein;
   
   + scripts: folder that contains the scripts used herein;

   + info: folder that contains some useful information (*i.e.*, papers, etc) used herein;
  
<br>

<br>

---

<br>

<br>

#### R packages used and respective versions

```{r References, message=FALSE, warning=FALSE, paged.print=FALSE}

## R packages and versions used in these analyses

sessionInfo()

```

<br>

<br>

---

<br>

<br>
<br>
<br>


## References

\printbibliography
