---
title: 'Report: Functional enrichment and plot DEG/markers upon integrated Cre3 and Lox2 scRNA samples'
author: "Jingtao Lilue and António Sousa - UBI-IGC (e-mail: <agsousa@igc.gulbenkian.pt>)"
date: "`r format(Sys.Date(), '%d/%m/%Y')`"
output: 
  html_document:
    toc: true 
    toc_float: true
    theme: united
    code_folding: hide
bibliography: bibliography.bib
---

<br>

---

<br>

<br>

## Request

<br>

This report includes a description and the results of the analyses requested by Elisa Jentho at 08/06/2021: 
   
   1 - Do functional enrichment analysis using only the functional database GO:BP. 
  
   2 - Re-do heatmaps of differential expression genes using log2 fold change values.

   3 - Plot additional markers in the integrated overlay UMAPs. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	fig.align = "center",
	message = FALSE,
	warning = FALSE
)
```

```{r klippy, echo=FALSE}
klippy::klippy(position = c('top', 'right'))
```

<br>

<br>

---

<br>

<br>

## About this report

<br>

The R version used was `r paste0(version$major, ".", version$minor)` [@r]. This report was built with the `rmarkdown` R package (v.`r packageVersion("rmarkdown")`) [@rmarkdown_1; @rmarkdown_2; @rmarkdown_3]. See the full list of packages and versions used at the end of this report - **R packages used and respective versions**.

The integrated 10x Lox2 and Cre3 samples processed previously were imported to plot the markers upon the UMAPs with the `Seurat` R package (v.`r packageVersion("Seurat")`) [@seurat_1; @seurat_2; @seurat_3; @seurat_4].

The figures and tables displayed through the report can be download in *pdf* and *tab-separated* format by clicking on the bottom left **Download plot** and **Download table** button that appears after each plot and table, respectively. 

```{r seed}

## Set seed 
set.seed(seed = 1024) # to keep reproducibility

```

```{r import packages, message=FALSE, warning=FALSE}

## Import packages 

library("dplyr", quietly = TRUE)
library("DT", quietly = TRUE) # package to print nice in the html report 
library("Seurat", quietly = TRUE)
library("gplots", quietly = TRUE)
library("Matrix", quietly = TRUE)
library("ggplot2", quietly = TRUE)
library("tidyr", quietly = TRUE)
library("readxl", quietly = TRUE)
library("gprofiler2", quietly = TRUE)
library("ComplexHeatmap", quietly = TRUE)

```

<br>

<br>

---

<br>

<br>

## Data analysis

<br>

### Import data

<br>

The **Cre3** and **Lox2** integrated samples were imported. 

```{r import seu}

## Import Seurat object with the re3 & Lox2 integrated samples

# import obj integrated
seu <- readRDS(file = "../results/int_28_05_21/R_objects/seu.rds")

```

It was also imported the differential gene expression tables obtained before. 

```{r import dge}

## Import DGE tables

dge_tables <- "../results/int_28_05_21/tables/dge_tables"
dge2import <- list.files(path = dge_tables, pattern = "clt_", full.names = TRUE)
names(dge2import) <- lapply(basename(dge2import), function(x) strsplit(x = x, split = "_")[[1]][1:2] %>% 
                              unlist(.) %>% paste(., collapse = "_")) %>% unlist(.)
dge <- lapply(dge2import, function(x) read.table(file = x, header = TRUE, sep = "\t", 
                                                 stringsAsFactors = FALSE))

```

It was also imported the markers to plot in the overlay integrated UMAPs as well as the markers for cluster 8 from the integrated Cre3 sample alone (`"PCHAS-130590"`, `"PCHAS-130630"`, `"PCHAS-130770"`, `"PCHAS-131600"`, `"PCHAS-134500"`). 

```{r import markers}

## Plot overlay markers for cluster 8 
markers_clt8 <- c("PCHAS-130590", "PCHAS-130630", "PCHAS-130770", "PCHAS-131600", "PCHAS-134500")

# import new markers
markers2import <- "../data/markers_to_plot_08_06_21/MArkers.xlsx"
markers_all <- read_excel(path = markers2import, col_names = FALSE)
colnames(markers_all) <- "Genes"

```

The new markers to plot are highlighted below. 

<br>

```{r echo=FALSE}
datatable(data = markers_all, extensions = "Buttons", 
          caption = "Table with the list of new markers to plot/highlight in the overlay UMAPs of the integrated Cre3 and Lox2 samples. Imported from the excel file provided: Markers_for_UMAP.xlsx.", 
          options = list(dom = "Blfrtip", buttons = c('csv', 'excel')))     
```

<br>

<br>

---

<br>

<br>

### Functional enrichment: GO:BP

<br>

In order to perform a functional enrichment analysis between the differentially expressed up- and down-regulated genes obtained from each pairwise comparison between each cluster across Cre3 *vs* Lox2, it was used the `gprofiler2` R package (v.`r packageVersion("gprofiler2")`) [@gprofiler2], an interface to the [g:Profiler](https://biit.cs.ut.ee/gprofiler/gost) web browser tool. The function `gost()` was applied in order to perform enrichment based on the list of up- or down-regulated genes, between each pairwise comparison (independently), against the annotated genes (domain_scope = "annotated") of the organism *Plasmodium chabaudi* (`organism = "pchabaudi"`). The gene lists were ordered by increasing adjusted *p*-value (`ordered_query = TRUE`) in order to generate a GSEA (Gene Set Enrichment Analysis) style *p*-values. This allows to start the enrichment testing from the top most biological relevant genes with subsequent tests involving larger sets of genes. In addition, only statistically significant (`user_threshold = 0.05`) enriched functions are returned (`significant = TRUE`) after multiple testing correction with the default method g:SCS (`correction_method = "g_SCS"`). Finally evidence codes are added to the final result (`evcodes = TRUE`). The functional enrichment analysis was performed only against the `GO:BP` functional database by providing the option `sources = "GO:BP"`.

<br>

It was tested functional enrichment for all clusters, with the exception of cluster 8, because it lacks differential gene expression data. It was given individually a list of up- (avg_log2FC > 0) or downregulated (avg_log2FC < 0) genes ranked (from the most significant to the lowest) by adjusted *p*-values (only genes with an adjusted *p*-value < 0.05). 

The result of this analysis is a plot and a table for each up- and down-regulated gene list from each cluster pairwise comparison across Cre3 *vs* Lox2, that shows the functional enrichment based on the -log10 of the adjusted *p*-value across different functional databases (see above). Only significant functions are reported. The table contains the following columns/information (the information below was copied entirely from the original vignette [here](https://cran.r-project.org/web/packages/gprofiler2/vignettes/gprofiler2.html#gene-list-functional-enrichment-analysis-with-gost)): 

<br>

   + **query**: the name of the input query which by default is the order of query with the prefix "query_". This can be changed by using a named list input.
   
   + **significant**: indicator for statistically significant results
   
   + **p_value**: hypergeometric p-value after correction for multiple testing
   
   + **term_size**: number of genes that are annotated to the term
    
   + **query_size**: number of genes that were included in the query. This might be different from the size of the original list if:
   
      + any genes were mapped to multiple Ensembl gene IDs
      
      + any genes failed to be mapped to Ensembl gene IDs
      
      + the parameter ordered_query = TRUE and the optimal cutoff for the term was found before the end of the query
        
      + the domain_scope was set to "annotated" or "custom"
    
   + **intersection_size**: the number of genes in the input query that are annotated to the corresponding term
    
   + **precision**: the proportion of genes in the input list that are annotated to the function (defined as intersection_size/query_size)
    
   + **recall**: the proportion of functionally annotated genes that the query recovers (defined as intersection_size/term_size)
    
   + **term_id**: unique term identifier (e.g GO:0005005)
   
   + **source**: the abbreviation of the data source for the term (e.g. GO:BP)
   
   + **term_name**: the short name of the function
   
   + **effective_domain_size**: the total number of genes "in the universe" used for the hypergeometric test
   
   + **source_order**: numeric order for the term within its data source (this is important for drawing the results)
   
   + **parents**: list of term IDs that are hierarchically directly above the term. For non-hierarchical data sources this points to an artificial root node.

<br>

In the bubble plots represented below, functions that have a -log10(adjusted *p*-value)>16 will be capped; *i.e.*, they will appear close to the vertical dashed black line. All these results reported below were run at 09/06/2021 using the archived version of the `gprofiler2` server - Ensembl 102, Ensembl Genomes 49 (database built on 2020-12-15): [https://biit.cs.ut.ee/gprofiler_archive3/e102_eg49_p15/gost](https://biit.cs.ut.ee/gprofiler_archive3/e102_eg49_p15/gost). 

```{r reg lists}

## Get up and down gene lists and parse them

reg_gene_list <- list()
for ( clt in names(dge) ) { # select dge by cluster and retrieve name
  sub_df <- dge[[clt]]
  #sub_df[,"Geneid"] <- rownames(sub_df)
  gene_ids_up <- sub_df %>% 
    filter(p_val_adj < 0.05 & avg_log2FC > 0) %>% 
    arrange(p_val_adj) %>% 
    pull(Geneid)
  gene_ids_down <- sub_df %>% 
    filter(p_val_adj < 0.05 & avg_log2FC < 0) %>% 
    arrange(p_val_adj) %>% 
    pull(Geneid)  
  reg_gene_list[[clt]] <- list()
  reg_gene_list[[clt]][["up"]] <- gsub(pattern = "PCHAS-", replacement = "PCHAS_", x = gene_ids_up)
  reg_gene_list[[clt]][["down"]] <- gsub(pattern = "PCHAS-", replacement = "PCHAS_", x = gene_ids_down)
}

# create folders.
r_object_folder <- "../results/func_enrich_plots_09_06_21/R_objects"
if ( ! dir.exists(r_object_folder) ) dir.create(r_object_folder, recursive = TRUE)
func_enrich_folder <- "../results/func_enrich_plots_09_06_21/tables/functional_enrichment"
if ( ! dir.exists(func_enrich_folder) ) dir.create(func_enrich_folder, recursive = TRUE)

```

```{r func enrich, echo=TRUE, eval=FALSE}

### Functional enrichment of DEG 

## Run gprofiler2
func_enrich <- list()
set_base_url("https://biit.cs.ut.ee/gprofiler_archive3/e102_eg49_p15")
for ( clt in names(reg_gene_list) ){ # loop over list and do functional enrichment
  #print(get_base_url())
  func_enrich[[clt]] <- list()
  set.seed(1024)
  func_enrich[[clt]][["up"]] <- gost(query = reg_gene_list[[clt]][["up"]], 
                                     organism = "pchabaudi", ordered_query = TRUE, 
                                     multi_query = FALSE, significant = TRUE, exclude_iea = FALSE, 
                                     measure_underrepresentation = FALSE, evcodes = TRUE, 
                                     user_threshold = 0.05, correction_method = "g_SCS", 
                                     domain_scope = "annotated", custom_bg = NULL, 
                                     numeric_ns = "", sources = "GO:BP")
  if ( ! is.null(func_enrich[[clt]][["up"]]$result) ) {
    sub_df_up <- func_enrich[[clt]][["up"]]$result %>% 
    apply(X = ., MARGIN = 2, FUN =  function(x) as.character(x)) 
    write.table(x = sub_df_up, 
                file = paste(func_enrich_folder, paste0(clt, "_up_functional_enrichment_table.tsv"), 
                             sep = "/"), 
                quote = FALSE, sep = "\t", row.names = FALSE, col.names = TRUE)
  }
  set.seed(1024)
  func_enrich[[clt]][["down"]] <- gost(query = reg_gene_list[[clt]][["down"]], 
                                       organism = "pchabaudi", ordered_query = TRUE, 
                                       multi_query = FALSE, significant = TRUE, exclude_iea = FALSE, 
                                       measure_underrepresentation = FALSE, evcodes = TRUE, 
                                       user_threshold = 0.05, correction_method = "g_SCS", 
                                       domain_scope = "annotated", custom_bg = NULL, 
                                       numeric_ns = "", sources = "GO:BP")
  if ( ! is.null(func_enrich[[clt]][["down"]]$result) ) {
    sub_df_down <- func_enrich[[clt]][["down"]]$result %>% 
    apply(X = ., MARGIN = 2, FUN =  function(x) as.character(x)) 
    write.table(x = sub_df_down, 
                file = paste(func_enrich_folder, paste0(clt, "_down_functional_enrichment_table.tsv"), 
                             sep = "/"), 
                quote = FALSE, sep = "\t", row.names = FALSE, col.names = TRUE)
  }
}

# Create R object
saveRDS(object = func_enrich, file = paste(r_object_folder, "func_enrich.rds", sep = "/"))

```

```{r import func enrich}

# Import functional enrichment object to avoid to have to run the API app
func_enrich <- readRDS(file = paste(r_object_folder, "func_enrich.rds", sep = "/"))

```

<br>

Among the clusters tested, it was only found significant functional enrichment pathways for cluster 3 (down), 4 (up) and 7 (up). 
```{r func enrich - barplots}

## Functional enrichments - barplots

# Create folder to save results
func_enrich_barplots_folder <- "../results/func_enrich_plots_09_06_21/plots/func_enrich_barplots"
if ( ! dir.exists(func_enrich_barplots_folder) ) dir.create(func_enrich_barplots_folder, recursive = TRUE)

# Plot and save data
func_enrich_list <- list()
for ( clt in names(func_enrich) ) { # loop over cluster
  if ( ! is.null(names(func_enrich[[clt]])) ) { # if cluster is not NULL
    func_enrich_list[[clt]] <- list()
    for ( reg in names(func_enrich[[clt]]) ) { # loop over regulation: 'up' and 'down'
      if ( reg == "up" ) {
        color <- "#E64B35FF"
      } else {
        color <- "#4DBBD5FF"
      }
      # Plot barplot of functional enrichment 
      func_enrich_list[[clt]][[reg]] <- func_enrich[[clt]][[reg]]$result %>% 
        arrange(source, p_value) %>%
        mutate(term_name = factor(term_name, levels = rev(unique(term_name)))) %>%
        ggplot(data = ., mapping = aes(x = term_name, y = -log10(p_value))) + #, fill = source)) + 
        geom_bar(stat = "identity", fill = color) + 
        #ggsci::scale_fill_npg(name = "Source") +
        #facet_grid( source ~ . , scales = "free_y", space = "free") +
        coord_flip() + 
        theme_bw() + 
        ylab("-log10 (adjusted p-value)") + 
        xlab("Term name") + 
        theme(axis.text = element_text(size = 10, color = "black"))
      # save
      ggsave(filename = paste(func_enrich_barplots_folder, paste0(clt, "_", reg, "_func_enrich_barplot.pdf"), sep = "/"), 
             plot = func_enrich_list[[clt]][[reg]], width = 6, height = 4)
    }
  }
}

```

<br>

<br>

   + **Cluster 3 (down)**
   
<br>

```{r}
print(func_enrich_list$clt_3$down)
```

<br>

```{r echo=FALSE}
## Download the latest plot
downloadthis::download_file(
  path = paste(func_enrich_barplots_folder, "clt_3_down_func_enrich_barplot.pdf", sep = "/"),
  output_name = "clt_3_down_func_enrich_barplot.pdf",
  button_label = "Download plot",
  button_type = "danger",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = TRUE
)
``` 

<br>

<br>

   + **Cluster 4 (up)**
   
<br>

```{r}
print(func_enrich_list$clt_4$up)
```

<br>

```{r echo=FALSE}
## Download the latest plot
downloadthis::download_file(
  path = paste(func_enrich_barplots_folder, "clt_4_up_func_enrich_barplot.pdf", sep = "/"),
  output_name = "clt_4_up_func_enrich_barplot.pdf",
  button_label = "Download plot",
  button_type = "danger",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = TRUE
)
``` 

<br>

<br>   
   
   
   + **Cluster 7 (up)**

<br>

```{r}
print(func_enrich_list$clt_7$up)
```

<br>

```{r echo=FALSE}
## Download the latest plot
downloadthis::download_file(
  path = paste(func_enrich_barplots_folder, "clt_7_up_func_enrich_barplot.pdf", sep = "/"),
  output_name = "clt_7_up_func_enrich_barplot.pdf",
  button_label = "Download plot",
  button_type = "danger",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = TRUE
)
``` 

<br>

<br>

---

<br>

<br>

### DGE: Cre3 vs Lox2 (heatmaps)

<br>

A gene was considered differentially expressed (and plotted) if it has an absolute log2FC>0 and an adjusted p-value<0.05. It was used the R package `ComplexHeatmap` (v.`r packageVersion("ComplexHeatmap")`) [@complexheatmap]. The same color gradient scale ranging from -4 to 4 was giving to all the heatmaps in order to make them comparable. For the heatmaps with more genes differentially expressed, only the top 10 up- and downregulated genes were labeled.  

<br>

   + **DGE heatmap: Cre3 vs Lox2 - cluster 0**

```{r dge - heatmaps - clt_0, fig.width=3, fig.height=6}

# path to save plots
dge_folder <- "../results/func_enrich_plots_09_06_21/plots/dge"
if ( ! dir.exists(dge_folder) ) dir.create(dge_folder)

## DGE - heatmaps - clt 0

# data
cluster <- "clt_0"
sub_df <- dge[[cluster]] %>% 
  filter(p_val_adj<0.05 & abs(avg_log2FC)>0) %>% 
  arrange(desc(avg_log2FC))
row.names(sub_df) <- sub_df$Geneid
sub_df <- sub_df[,"avg_log2FC", drop = FALSE]
row_annot <- rowAnnotation(foo = anno_mark(at = c(1:2,45:54), 
                                           labels = row.names(sub_df)[c(1:2,45:54)], 
                                           labels_gp = gpar(fontsize = 8.5)))
col_fun <- circlize::colorRamp2(c(-4, 0, 4), c("blue", "white", "red"))

# save
set.seed(1024)
heat_clt0 <- Heatmap(matrix = as.matrix(sub_df), name = "Average\nlog2FC", 
                     cluster_columns = FALSE, cluster_rows = FALSE, 
                     show_column_names = FALSE, col = col_fun, 
                     show_row_names = FALSE,
                     right_annotation = row_annot)
set.seed(1024)
pdf(file = paste(dge_folder, paste0(cluster, "_dge_heatmap.pdf"), sep = "/"), 
    width = 3, height = 0.1 * nrow(sub_df))
print(heat_clt0)
dev.off()

# print
set.seed(1024)
print(heat_clt0)

```

<br>

```{r echo=FALSE}
## Download the latest plot
downloadthis::download_file(
  path = paste(dge_folder, paste0(cluster, "_dge_heatmap.pdf"), sep = "/"),
  output_name = paste0(cluster, "_dge_heatmap.pdf"),
  button_label = "Download plot",
  button_type = "danger",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = TRUE
)
``` 

<br>

<br>

   + **DGE heatmap: Cre3 vs Lox2 - cluster 1**

```{r dge - heatmaps - clt_1, fig.width=3, fig.height=6}

## DGE - heatmaps - clt 1

# data
cluster <- "clt_1"
sub_df <- dge[[cluster]] %>% 
  filter(p_val_adj<0.05 & abs(avg_log2FC)>0) %>% 
  arrange(desc(avg_log2FC))
row.names(sub_df) <- sub_df$Geneid
sub_df <- sub_df[,"avg_log2FC", drop = FALSE]
row_annot <- rowAnnotation(foo = anno_mark(at = c(1:10,44:53), 
                                           labels = row.names(sub_df)[c(1:10,44:53)], 
                                           labels_gp = gpar(fontsize = 8.5)))
col_fun <- circlize::colorRamp2(c(-4, 0, 4), c("blue", "white", "red"))

# save
set.seed(1024)
heat_clt1 <- Heatmap(matrix = as.matrix(sub_df), name = "Average\nlog2FC", 
                     cluster_columns = FALSE, cluster_rows = FALSE, 
                     show_column_names = FALSE, col = col_fun, 
                     show_row_names = FALSE,
                     right_annotation = row_annot)
set.seed(1024)
pdf(file = paste(dge_folder, paste0(cluster, "_dge_heatmap.pdf"), sep = "/"), 
    width = 3, height = 0.1 * nrow(sub_df))
print(heat_clt1)
dev.off()

# print
set.seed(1024)
print(heat_clt1)

```

<br>

```{r echo=FALSE}
## Download the latest plot
downloadthis::download_file(
  path = paste(dge_folder, paste0(cluster, "_dge_heatmap.pdf"), sep = "/"),
  output_name = paste0(cluster, "_dge_heatmap.pdf"),
  button_label = "Download plot",
  button_type = "danger",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = TRUE
)
``` 

<br>

<br>

   + **DGE heatmap: Cre3 vs Lox2 - cluster 2**

```{r dge - heatmaps - clt_2, fig.width=3, fig.height=6}

## DGE - heatmaps - clt 2

# data
cluster <- "clt_2"
sub_df <- dge[[cluster]] %>% 
  filter(p_val_adj<0.05 & abs(avg_log2FC)>0) %>% 
  arrange(desc(avg_log2FC))
row.names(sub_df) <- sub_df$Geneid
sub_df <- sub_df[,"avg_log2FC", drop = FALSE]
row_annot <- rowAnnotation(foo = anno_mark(at = c(1:9,55:64), 
                                           labels = row.names(sub_df)[c(1:9,55:64)], 
                                           labels_gp = gpar(fontsize = 8.5)))
col_fun <- circlize::colorRamp2(c(-4, 0, 4), c("blue", "white", "red"))

# save
set.seed(1024)
heat_clt2 <- Heatmap(matrix = as.matrix(sub_df), name = "Average\nlog2FC", 
                     cluster_columns = FALSE, cluster_rows = FALSE, 
                     show_column_names = FALSE, col = col_fun, 
                     show_row_names = FALSE,
                     right_annotation = row_annot)
set.seed(1024)
pdf(file = paste(dge_folder, paste0(cluster, "_dge_heatmap.pdf"), sep = "/"), 
    width = 3, height = 0.1 * nrow(sub_df))
print(heat_clt2)
dev.off()

# print
set.seed(1024)
print(heat_clt2)

```

<br>

```{r echo=FALSE}
## Download the latest plot
downloadthis::download_file(
  path = paste(dge_folder, paste0(cluster, "_dge_heatmap.pdf"), sep = "/"),
  output_name = paste0(cluster, "_dge_heatmap.pdf"),
  button_label = "Download plot",
  button_type = "danger",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = TRUE
)
``` 

<br>

<br>

   + **DGE heatmap: Cre3 vs Lox2 - cluster 3**

```{r dge - heatmaps - clt_3, fig.width=3, fig.height=6}

## DGE - heatmaps - clt 3

# data
cluster <- "clt_3"
sub_df <- dge[[cluster]] %>% 
  filter(p_val_adj<0.05 & abs(avg_log2FC)>0) %>% 
  arrange(desc(avg_log2FC))
row.names(sub_df) <- sub_df$Geneid
sub_df <- sub_df[,"avg_log2FC", drop = FALSE]
row_annot <- rowAnnotation(foo = anno_mark(at = c(1:10,196:205), 
                                           labels = row.names(sub_df)[c(1:10,196:205)], 
                                           labels_gp = gpar(fontsize = 8.5)))
col_fun <- circlize::colorRamp2(c(-4, 0, 4), c("blue", "white", "red"))

# save
set.seed(1024)
heat_clt3 <- Heatmap(matrix = as.matrix(sub_df), name = "Average\nlog2FC", 
                     cluster_columns = FALSE, cluster_rows = FALSE, 
                     show_column_names = FALSE, col = col_fun, 
                     show_row_names = FALSE,
                     right_annotation = row_annot)
set.seed(1024)
pdf(file = paste(dge_folder, paste0(cluster, "_dge_heatmap.pdf"), sep = "/"), 
    width = 3, height = 0.1 * nrow(sub_df))
print(heat_clt3)
dev.off()

# print
set.seed(1024)
print(heat_clt3)

```

<br>

```{r echo=FALSE}
## Download the latest plot
downloadthis::download_file(
  path = paste(dge_folder, paste0(cluster, "_dge_heatmap.pdf"), sep = "/"),
  output_name = paste0(cluster, "_dge_heatmap.pdf"),
  button_label = "Download plot",
  button_type = "danger",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = TRUE
)
```

<br>

<br>

   + **DGE heatmap: Cre3 vs Lox2 - cluster 4**

```{r dge - heatmaps - clt_4, fig.width=3, fig.height=6}

## DGE - heatmaps - clt 4

# data
cluster <- "clt_4"
sub_df <- dge[[cluster]] %>% 
  filter(p_val_adj<0.05 & abs(avg_log2FC)>0) %>% 
  arrange(desc(avg_log2FC))
row.names(sub_df) <- sub_df$Geneid
sub_df <- sub_df[,"avg_log2FC", drop = FALSE]
row_annot <- rowAnnotation(foo = anno_mark(at = c(1:10,84:93), 
                                           labels = row.names(sub_df)[c(1:10,84:93)], 
                                           labels_gp = gpar(fontsize = 8.5)))
col_fun <- circlize::colorRamp2(c(-4, 0, 4), c("blue", "white", "red"))

# save
set.seed(1024)
heat_clt4 <- Heatmap(matrix = as.matrix(sub_df), name = "Average\nlog2FC", 
                     cluster_columns = FALSE, cluster_rows = FALSE, 
                     show_column_names = FALSE, col = col_fun, 
                     show_row_names = FALSE,
                     right_annotation = row_annot)
set.seed(1024)
pdf(file = paste(dge_folder, paste0(cluster, "_dge_heatmap.pdf"), sep = "/"), 
    width = 3, height = 0.1 * nrow(sub_df))
print(heat_clt4)
dev.off()

# print
set.seed(1024)
print(heat_clt4)

```

<br>

```{r echo=FALSE}
## Download the latest plot
downloadthis::download_file(
  path = paste(dge_folder, paste0(cluster, "_dge_heatmap.pdf"), sep = "/"),
  output_name = paste0(cluster, "_dge_heatmap.pdf"),
  button_label = "Download plot",
  button_type = "danger",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = TRUE
)
```

<br>

<br>

   + **DGE heatmap: Cre3 vs Lox2 - cluster 5**

```{r dge - heatmaps - clt_5, fig.width=3, fig.height=6}

## DGE - heatmaps - clt 5

# data
cluster <- "clt_5"
sub_df <- dge[[cluster]] %>% 
  filter(p_val_adj<0.05 & abs(avg_log2FC)>0) %>% 
  arrange(desc(avg_log2FC))
row.names(sub_df) <- sub_df$Geneid
sub_df <- sub_df[,"avg_log2FC", drop = FALSE]
# row_annot <- rowAnnotation(foo = anno_mark(at = c(1:6,9:18), 
#                                            labels = row.names(sub_df)[c(1:6,9:18)], 
#                                            labels_gp = gpar(fontsize = 8.5)))
col_fun <- circlize::colorRamp2(c(-4, 0, 4), c("blue", "white", "red"))

# save
set.seed(1024)
heat_clt5 <- Heatmap(matrix = as.matrix(sub_df), name = "Average\nlog2FC", 
                     cluster_columns = FALSE, cluster_rows = FALSE, 
                     show_column_names = FALSE, col = col_fun, 
                     show_row_names = TRUE, row_names_gp = gpar(fontsize = 6.5))#,
                     #right_annotation = row_annot)
set.seed(1024)
pdf(file = paste(dge_folder, paste0(cluster, "_dge_heatmap.pdf"), sep = "/"), 
    width = 3, height = 0.1 * nrow(sub_df))
print(heat_clt5)
dev.off()

# print
set.seed(1024)
print(heat_clt5)

```

<br>

```{r echo=FALSE}
## Download the latest plot
downloadthis::download_file(
  path = paste(dge_folder, paste0(cluster, "_dge_heatmap.pdf"), sep = "/"),
  output_name = paste0(cluster, "_dge_heatmap.pdf"),
  button_label = "Download plot",
  button_type = "danger",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = TRUE
)
```

<br>

<br>

   + **DGE heatmap: Cre3 vs Lox2 - cluster 6**

```{r dge - heatmaps - clt_6, fig.width=3, fig.height=6}

## DGE - heatmaps - clt 6

# data
cluster <- "clt_6"
sub_df <- dge[[cluster]] %>% 
  filter(p_val_adj<0.05 & abs(avg_log2FC)>0) %>% 
  arrange(desc(avg_log2FC))
row.names(sub_df) <- sub_df$Geneid
sub_df <- sub_df[,"avg_log2FC", drop = FALSE]
# row_annot <- rowAnnotation(foo = anno_mark(at = c(1:6,9:18), 
#                                            labels = row.names(sub_df)[c(1:6,9:18)], 
#                                            labels_gp = gpar(fontsize = 8.5)))
col_fun <- circlize::colorRamp2(c(-4, 0, 4), c("blue", "white", "red"))

# save
set.seed(1024)
heat_clt6 <- Heatmap(matrix = as.matrix(sub_df), name = "Average\nlog2FC", 
                     cluster_columns = FALSE, cluster_rows = FALSE, 
                     show_column_names = FALSE, col = col_fun, 
                     show_row_names = TRUE, row_names_gp = gpar(fontsize = 6.5))#,
                     #right_annotation = row_annot)
set.seed(1024)
pdf(file = paste(dge_folder, paste0(cluster, "_dge_heatmap.pdf"), sep = "/"), 
    width = 3, height = 0.1 * nrow(sub_df))
print(heat_clt6)
dev.off()

# print
set.seed(1024)
print(heat_clt6)

```

<br>

```{r echo=FALSE}
## Download the latest plot
downloadthis::download_file(
  path = paste(dge_folder, paste0(cluster, "_dge_heatmap.pdf"), sep = "/"),
  output_name = paste0(cluster, "_dge_heatmap.pdf"),
  button_label = "Download plot",
  button_type = "danger",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = TRUE
)
```

<br>

<br>

   + **DGE heatmap: Cre3 vs Lox2 - cluster 7**

```{r dge - heatmaps - clt_7, fig.width=3, fig.height=6}

## DGE - heatmaps - clt 7

# data
cluster <- "clt_7"
sub_df <- dge[[cluster]] %>% 
  filter(p_val_adj<0.05 & abs(avg_log2FC)>0) %>% 
  arrange(desc(avg_log2FC))
row.names(sub_df) <- sub_df$Geneid
sub_df <- sub_df[,"avg_log2FC", drop = FALSE]
row_annot <- rowAnnotation(foo = anno_mark(at = c(1:10,48:57),
                                           labels = row.names(sub_df)[c(1:10,48:57)],
                                           labels_gp = gpar(fontsize = 8.5)))
col_fun <- circlize::colorRamp2(c(-4, 0, 4), c("blue", "white", "red"))

# save
set.seed(1024)
heat_clt7 <- Heatmap(matrix = as.matrix(sub_df), name = "Average\nlog2FC", 
                     cluster_columns = FALSE, cluster_rows = FALSE, 
                     show_column_names = FALSE, col = col_fun, 
                     show_row_names = FALSE, 
                     right_annotation = row_annot)
set.seed(1024)
pdf(file = paste(dge_folder, paste0(cluster, "_dge_heatmap.pdf"), sep = "/"), 
    width = 3, height = 0.1 * nrow(sub_df))
print(heat_clt7)
dev.off()

# print
set.seed(1024)
print(heat_clt7)

```

<br>

```{r echo=FALSE}
## Download the latest plot
downloadthis::download_file(
  path = paste(dge_folder, paste0(cluster, "_dge_heatmap.pdf"), sep = "/"),
  output_name = paste0(cluster, "_dge_heatmap.pdf"),
  button_label = "Download plot",
  button_type = "danger",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = TRUE
)
```

<br>

<br>

---

<br>

<br>

### Plot markers over overlay UMAPs

<br>

To plot markers it was used the integrated `Seurat` object with the Cre3 and Lox2 samples. The assay used was `RNA` and the slot `data` with the counts normalized with the `LogNormalize` method ([read more](https://satijalab.org/seurat/reference/normalizedata)). 

The individual plots of the new markers highlighted through the integrated UMAPs can be found at: `results/func_enrich_plots_09_06_21/plots/markers_all`.  

```{r plot new markers}

## Plot new markers

# folder to save plots
plot_mkr_folder <- "../results/func_enrich_plots_09_06_21/plots/markers_all"
if ( ! dir.exists(plot_mkr_folder) ) dir.create(plot_mkr_folder)

# check assay 
stopifnot(DefaultAssay(seu) == "RNA")
marker_plots <- list()
for ( mkr in markers_all$Genes ) { # loop over marker genes within a cluster
  stopifnot( mkr %in% row.names(seu) ) # stop if gene does not exist in the Seurat obj
  marker_plots[[mkr]] <- FeaturePlot(object = seu, features = mkr, 
                                     slot = "data", reduction = "umap", 
                                     pt.size = 0.001)
  ggsave(filename = paste(plot_mkr_folder, paste0(mkr, "_gene_UMAP.pdf"), sep = "/"), 
         plot = marker_plots[[mkr]], height = 5, width = 5)
}

```

```{r fig.width=10, fig.height=9}

cowplot::plot_grid(plotlist = marker_plots[1:9], ncol = 3)

```

```{r fig.width=10, fig.height=9}

cowplot::plot_grid(plotlist = marker_plots[10:18], ncol = 3)

```

```{r fig.width=10, fig.height=9}

cowplot::plot_grid(plotlist = marker_plots[19:27], ncol = 3)

```

<br>

<br>

### Plot markers for cluster 8 by Lox2 UMAP

<br>

To plot cluster 8 markers it was used the integrated `Seurat` object with the Cre3 sample only. The assay used was `RNA` and the slot `data` with the counts normalized with the `LogNormalize` method ([read more](https://satijalab.org/seurat/reference/normalizedata)). 

The individual plots of the new markers highlighted through the integrated UMAPs can be found at: `results/func_enrich_plots_09_06_21/plots/markers_clt8`.  

```{r mkr clt 8 lox2, fig.width=10, fig.height=6}

## Plot markers for cluster 8 by Lox2 UMAP

# folder to save plots
plot_mkr_clt8_folder <- "../results/func_enrich_plots_09_06_21/plots/markers_clt8"
if ( ! dir.exists(plot_mkr_clt8_folder) ) dir.create(plot_mkr_clt8_folder)

# get Seurat Lox2 obj
cre3 <- SplitObject(object = seu, split.by = "orig.ident")
cre3 <- cre3$Cre3

# save plots
marker_plot_clt8 <- list()
for ( mkr in markers_clt8 ) {
  stopifnot( mkr %in% row.names(cre3) ) # stop if gene does not exist in the Seurat obj
  marker_plot_clt8[[mkr]] <- FeaturePlot(object = cre3, features = mkr, 
                                         slot = "data", reduction = "umap", 
                                         pt.size = 0.001)
  ggsave(filename = paste(plot_mkr_clt8_folder, paste0(mkr, "_gene_UMAP.pdf"), sep = "/"), 
         plot = marker_plot_clt8[[mkr]], height = 5, width = 5)
  
}

# plot all
cowplot::plot_grid(plotlist = marker_plot_clt8, ncol = 3)

```

<br>

<br>

---

<br>

<br>

### Orthology conversion: P. chabaudi *vs* P. falciparum 3D7

<br>

Orthology search was done with the `gprofiler2` R package (v.`r packageVersion("gprofiler2")`) [@gprofiler2], an interface to the [g:Profiler](https://biit.cs.ut.ee/gprofiler/gost) web browser tool. The function `gorth()` was applied in order get orthologous genes of *Plasmodium chabaudi chabaudi* (v.PCHAS01; `source_organism = "pchabaudi"`) against *Plasmodium falciparum 3D7* (v.ASM276v2; `target_organism = "pfalciparum"`). The query gene list of *P. chabaudi* genes consisted in all the genes that are expressed in the `Seurat` integrated object (n=`r length(row.names(seu))`). Genes without correspondence were filtered out (`filter_na = TRUE`). One `*P. chabaudi` may have more than one orthologous gene (`mthreshold = Inf`). This conversion was done at 09/06/2021 using the archived version of the `gprofiler2` server - Ensembl 102, Ensembl Genomes 49 (database built on 2020-12-15): [https://biit.cs.ut.ee/gprofiler_archive3/e102_eg49_p15/gost](https://biit.cs.ut.ee/gprofiler_archive3/e102_eg49_p15/gost).    

```{r orthology conversion}

## Convert orthologous genes between P. chabaudi *vs* P. falciparum 3D7

genes_exp <- row.names(seu)
genes_exp <- gsub(pattern = "PCHAS-", replacement = "PCHAS_", x = genes_exp)
set.seed(1024)
set_base_url("https://biit.cs.ut.ee/gprofiler_archive3/e102_eg49_p15")
orth_pchabaudi_genes <- gorth(query = genes_exp, source_organism = "pchabaudi", target_organism = "pfalciparum")
write.table(file = "../results/func_enrich_plots_09_06_21/tables/orthologous_genes_Pchabaudi_vs_Pfalciparum3D7.tsv", 
            x = orth_pchabaudi_genes, sep = "\t", row.names = FALSE, quote = FALSE)

```

Among the *P. chabaudi* `r length(row.names(seu))` genes queried, it was only found `r nrow(orth_pchabaudi_genes)` hits presented below. 

<br>

```{r echo=FALSE}
datatable(data = orth_pchabaudi_genes, extensions = "Buttons", 
          caption = "Table with the list of orthologous genes between Plasmodium chabaudi chabaudi and Plasmodium falciparum 3D7 obtained with gprofiler2.", 
          options = list(dom = "Blfrtip", buttons = c('csv', 'excel')))     
```

<br>

```{r echo=FALSE}
## Download the latest table
downloadthis::download_file(
  path = "../results/func_enrich_plots_09_06_21/tables/orthologous_genes_Pchabaudi_vs_Pfalciparum3D7.tsv",
  output_name = "orthologous_genes_Pchabaudi_vs_Pfalciparum3D7.tsv",
  button_label = "Download table",
  button_type = "danger",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = TRUE
)
``` 

<br>

<br>

---

<br>

<br>
    
## Deliver

Folders inside the project folder:

   + results: folder that contains all the results obtained in this analysis;

   + report: folder that contains the report and code used herein;

   + data: folder that contains the data used herein;
   
   + scripts: folder that contains the scripts used herein;

   + info: folder that contains some useful information (*i.e.*, papers, etc) used herein;
  
<br>

<br>

---

<br>

<br>

#### R packages used and respective versions

```{r References, message=FALSE, warning=FALSE, paged.print=FALSE}

## R packages and versions used in these analyses

sessionInfo()

```

<br>

<br>

---

<br>

<br>
<br>
<br>


## References

\printbibliography
